Network based multifactorial modelling microRNA-target interaction.

### 2019-08-09 article.Rmd
overview

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs (TOFIX: Hocam etkileşim faktörlerinden dolayı ekspresyonları yakın olsa bile baskılanma aktitesinin aynı sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?). We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package at https://github.com/selcenari/ceRNAnetsim. (TODO the link will change)

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood, miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’end bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that binding location could indicate functionality of miRNA:target interaction or be effective on abundance of targets(TOFIX abundance etkisi derken, bazen azaltabilir, bazen arttırabilir, manasında mı?, evet hocam anlaşılmıyorsa değiştirebilirim). It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’ end of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

[#TODO]: <> "there are many et. al and et. al. , with or without dot(.) please make sure the formatting is correct and consistent"

@@

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger. c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity](Fig1.jpg){#fig1}

[#TODO]: <> "caption needs more text, every color should be explained"

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$) (#TODO isn't $C_g$ gene expressions in the group is equal to total of $C_{gi}$ in group?). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

\begin{equation} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Factors effecting binding determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Factors effecting efficiency determine proportion of target mRNA degraded by bound miRNA (#TODO fix this sentence). In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Efficiency factors determine how many of miRNA:target complexes will result in inhibition and binding region on target drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{4}\label{eq:4}
\end{equation}
\begin{equation}
Eff_{gi}= Eff_{gi}*RE^\prime_{gi} \tag{5}\label{eq:5}
\end{equation}

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from TCGA(#TODO cite). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#TODO name of database and its citation). Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (#TODO last sentence needs some clarification, after "but cooperative...") (\autoref{fig1}d). Genes targeted by multiple miRNAs act as a trigger for adjacent local neighborhood of targeting miRNAs, causing changes in expression levels of genes outside the local neighborhood of original trigger gene. Therefore, primary expression level change in gene (G2) causes changes in other group of genes (G5 and G6) even though original trigger gene (G2) and genes in other group are not targeted by common miRNA. In addition, as shown by ceRNA hypothesis model of Ala et al., after the increase of gene expression level of G2, the miRNA that is found in the same group (M1) tended to be more (#TODO isn't it less?) repressive on its remaining targets (G1, G3 and G4). It's important to note that the changes in gene expression levels will have more pronounced effect if miRNA:target ratio is high, i.e., more miRNA available per target, which was reported in previous findings [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. [][#TODO]

@@

### Multifactorial networks

[#TODO]: <> "title should be more informative"

Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in  Figure \autoref{fig1} is recalculated with additional factors in effect [][#TODO] with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed (compare miRNA distribution values between Figure \autoref{fig1}a and Figure \autoref{fig3}a). When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. [][#TODO2]

@@

[#TODO]: <> "didnt we perform upregulation of Gene2, why did we say Gene4 here?"
[#TODO2]: <> "Removed these sentences because it was somewhat confusing: In our approach, the elements behave differently based on type (competing or miRNA). When a competing element was changed, whole elements out of miRNAs are re-regulated. But, if a miRNA was perturbed, whole competing elements are re-regulated and miRNA preserves expression at after perturbation."

>>>>

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs (answer: Hocam etkileşim faktörlerinden dolayı ekspresyonları yakın olsa bile baskılanma aktitesinin aynı sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?). We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood, miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’end bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that binding location could indicate functionality of miRNA:target interaction or be effective on abundance of targets(TOFIX abundance etkisi derken, bazen azaltabilir, bazen arttırabilir, manasında mı? answer: evet hocam anlaşılmıyorsa değiştirebilirim). It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’ end of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

@@

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown with red in figure). c) Distribution of miRNA1 (M1) changes. d) The change at expression (shown with light red because minor changes are occured on their expression values.) of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity](Fig1.jpg){#fig1}

[#TODO]: <> "caption needs more text, every color should be explained"
[answer]: <> "explained in caption."

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$) (#TODO isn't $C_g$ gene expressions in the group is equal to total of $C_{gi}$ in group? answer: its my false. now its corect.). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

\begin{equation} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_{gi} \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Factors effecting binding determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Degradation efficiency factors assign amount of degradated target amount in miRNA:target pairs.(#TODO fix this sentence: answer:). In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Efficiency factors determine how many of miRNA:target complexes will result in inhibition and binding region on target drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{4}\label{eq:4}
\end{equation}
\begin{equation}
Eff_{gi}= Eff_{gi}*RE^\prime_{gi} \tag{5}\label{eq:5}
\end{equation}

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from TCGA(#TODO cite). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#TODO name of database and its citation)->suplementary datasets of [@helwak_mapping_2013; @moore_mirnatarget_2015] . Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (#TODO last sentence needs some clarification, after "but cooperative...") (\autoref{fig1}d). Genes targeted by multiple miRNAs act as a trigger for adjacent local neighborhood of targeting miRNAs, causing changes in expression levels of genes outside the local neighborhood of original trigger gene. Therefore, primary expression level change in gene (G2) causes changes in other group of genes (G5 and G6) even though original trigger gene (G2) and genes in other group are not targeted by common miRNA. In addition, as shown by ceRNA hypothesis model of Ala et al., after the increase of gene expression level of G2, the miRNA that is found in the same group (M1) tended to be less repressive on its remaining targets (G1, G3 and G4). It's important to note that the changes in gene expression levels will have more pronounced effect if miRNA:target ratio is high, i.e., more miRNA available per target, which was reported in previous findings [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. [][#TODO]

@@

### Competing Endogenous RNA:Target networks based on interaction factors

[#TODO]: <> "title should be more informative"-> answer.

Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in  Figure \autoref{fig1} is recalculated with additional factors in effect [][#TODO] with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed (compare miRNA distribution values between Figure \autoref{fig1}a and Figure \autoref{fig3}a). When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. [][#TODO2]

answer: knitr::kable(minsamp): figure ü fig1_2 dosyasında chunk2ye attım. Burada bir hata veriyor. Düzenleyemedim.

@@

[#TODO]: <> "didnt we perform upregulation of Gene2, why did we say Gene4 here?" 
[answer]: <> "Because, in fact, We want to show that triggering element is significant in the same dataset, in the same conditions. Maybe I must add a conjuction here."
